Language learning in context via recursive pragmatic reasoning

### 2013-05-31 praglearn.tex
intro tweak

The goal of our current work is to investigate the possibilities for
integrating models of recursive pragmatic reasoning with models of
language learning, with the hope of capturing phenomena in both
domains. We begin by describing previous models in each domain as well
as proposals for bringing the two together. We next simulate findings
on pragmatic inference in one-shot games (replicating previous
work). We then build on these results to simulate the results of pragmatic
learning in the language acquisition setting where one communicator is
uncertain about the lexicon and in iterated communication games where both
communicators are uncertain about the lexicon. We end by discussing the implications of our results for
work on language change.

@@

This resolves one paradox, but as soon as we attempt to add uncertainty about the meanings of words
to such a model, new paradoxes arise. Suppose the listener is a young
child who is uncertain about the lexicon their partner is using. The
obvious solution is for them to place a prior on the lexicon; they
then update their posterior based on whatever utterances and
contextual cues they observe, and in the mean time interpret each
utterance by making their best guess, marginalizing out this
uncertainty. This basic structure is captured in previous models of
Bayesian word learning \cite{frank2009}. But when combined with the
recursive pragmatic model, a new question arises: Given such a
listener, what model should the speaker use? A rational speaker
attempts to maximize the listener's likelihood of understanding, so if
an uncertain listener interpets by marginalizing over some posterior,
then a knowledgeable speaker should calculate this posterior and
marginalize themselves. But then the speaker is not using any
particular lexicon at all, and a rational learning listener should
therefore not be trying to infer the speaker's lexicon.\footnote{Of
  course, in reality both parties will generally have some
  uncertainty, making the situation even worse. If we start from an
  uncertain listener with a prior over lexicons, then a first-level
  uncertain speaker needs a prior over priors on lexicons, a
  second-level uncertain listener needs a prior over priors over
  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
  recursion was bad enough, but at least each step had a constant
  cost. This new recursion produces hyper$^n$-distributions for which
  inference almost immediately becomes intractable even in principle,
  since the dimensionality of the learning problem increases with each
  step. Yet, without this addition of new uncertainty at each level,
  the model would dissolve back into certainty as in the previous
  paragraph, making learning
  impossible.}
% In fact, such an uncertain listener's uncertainty dissolves; they
% may not know the lexicon, but they do
% know their posterior over the lexicon -- and because
% this is all that the speaker cares about, there is
% nothing for the listener to learn from the speaker's
% behavior.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes
their utility, with utility now being defined in terms of the
informativity of the expectation (over lexicons) that the listener
will have for the right referent\footnote{An alternative model would
  have the speaker take the expectation over informativity, instead of
  the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating {\sc all} of them, the listener is lead
to believe that {\sc some-but-not-all} is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.

>>>>

The goal of our current work is to investigate the possibilities for
integrating models of recursive pragmatic reasoning with models of
language learning, with the hope of capturing phenomena in both
domains. We begin by describing a proposal for bringing the two together, noting several issues in applying a purely normative framework to recursive reasoning under uncertainty. We next simulate findings
on pragmatic inference in one-shot games (replicating previous
work). We then build on these results to simulate the results of pragmatic
learning in the language acquisition setting where one communicator is
uncertain about the lexicon and in iterated communication games where both
communicators are uncertain about the lexicon.

@@

As soon as we attempt to add uncertainty about the meanings of words
to such a model, new paradoxes arise. Suppose the listener is a young
child who is uncertain about the lexicon their partner is using. The
obvious solution is for them to place a prior on the lexicon; they
then update their posterior based on whatever utterances and
contextual cues they observe, and in the mean time interpret each
utterance by making their best guess, marginalizing out this
uncertainty. This basic structure is captured in previous models of
Bayesian word learning \cite{frank2009}. But when combined with the
recursive pragmatic model, a new question arises: Given such a
listener, what model should the speaker use? A rational speaker
attempts to maximize the listener's likelihood of understanding, so if
an uncertain listener interpets by marginalizing over some posterior,
then a knowledgeable speaker should calculate this posterior and
marginalize themselves. But then the speaker is not using any
particular lexicon at all, and a rational learning listener should
therefore not be trying to infer the speaker's lexicon.\footnote{Of
  course, in reality both parties will generally have some
  uncertainty, making the situation even worse. If we start from an
  uncertain listener with a prior over lexicons, then a first-level
  uncertain speaker needs a prior over priors on lexicons, a
  second-level uncertain listener needs a prior over priors over
  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
  recursion was bad enough, but at least each step had a constant
  cost. This new recursion produces hyper$^n$-distributions for which
  inference almost immediately becomes intractable even in principle,
  since the dimensionality of the learning problem increases with each
  step. Yet, without this addition of new uncertainty at each level,
  the model would dissolve back into certainty as in the previous
  paragraph, making learning
  impossible.}
% In fact, such an uncertain listener's uncertainty dissolves; they
% may not know the lexicon, but they do
% know their posterior over the lexicon -- and because
% this is all that the speaker cares about, there is
% nothing for the listener to learn from the speaker's
% behavior.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating {\sc all} of them, the listener is lead
to believe that {\sc some-but-not-all}' is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.
